---
title: "Iris LDA and QDA"
author: "Data Mining"
output: pdf_document
---

## The iris  data

The iris data set is a classical data set first studied by R. A. Fisher in 1936. It contains measurements of sepal and petal length and width for three different species of the iris plant. In this exercise we are concerned with the prediction of species from sepal measurements.

We first load the data. We rename the species 1, 2, and 3 for simplicity, and define a new data frame *niris* with columns containing the species *Species*, sepal length *Slength*, and sepal width *Swidth*. 

library(neuralnet)
library(MASS)

```{r}
data(iris)
X = iris[,1:2]
y = rep(0,nrow(iris)) #Categories
y[iris[,5]=="setosa"] = 1
y[iris[,5]=="versicolor"] = 2
y[iris[,5]=="virginica"] = 3

niris = data.frame(Species=y, SLength = X[,1], SWidth = X[,2])

niris2 = data.frame(SLength = X[,1], SWidth = X[,2])
niris2$setosa     <- as.integer(iris[,5] == "setosa")
niris2$versicolor <- as.integer(iris[,5] == "versicolor")
niris2$virginica  <- as.integer(iris[,5] == "virginica")

```


## Classification error

- Randomly split the data set in two data sets of equal size and name them *train* and *test*.
- Fit an LDA and a QDA model on the data set *train*.
- Use *predict()* to predict the species of the observations in the data set *test*.
- How many observations in the *test* data set are misclassified by the two methods?

## Classification regions

- Plot the classification regions using LDA and QDA on the whole data set *niris* by modifying the code below.

First fit LDA and QDA models by editing the following code:

```{r}
# Data Split
set.seed(55)
index = sample(1:nrow(niris), round(0.5 * nrow(niris)), replace=FALSE)
train_iris = niris[index, ]
test_iris = niris[-index, ]
test_iris$Species  <- factor(test_iris$Species,  levels=1:3, labels=c("setosa","versicolor","virginica"))
train_iris$Species <- factor(train_iris$Species, levels=1:3, labels=c("setosa","versicolor","virginica"))
index2 = sample(1:nrow(niris2), round(0.5 * nrow(niris2)), replace = FALSE)
train_iris2 = niris2[index2, ]
test_iris2 = niris2[-index2, ]

# Models
nn = neuralnet("setosa + versicolor + virginica ~ SLength + SWidth", data=train_iris2, hidden=2, linear.output = FALSE, act.fct = "logistic")
lda.fit = lda(Species ~ SLength + SWidth, data=train_iris)
qda.fit = qda(Species ~ SLength + SWidth, data=train_iris)
plot(nn)

pr.nn = compute(nn, test_iris2[, c("SLength", "SWidth")])
pr.lda = predict(lda.fit, test_iris)
pr.qda = predict(qda.fit, test_iris)

pr.nn.mat = pr.nn$net.result
colnames(pr.nn.mat) = c("setosa", "versicolor", "virginica")

pr.nn.class = apply(pr.nn.mat, 1, which.max)
pr.nn.class = factor(pr.nn.class, levels=1:3, labels = c("setosa", "versicolor", "virginica"))

# Confusion matrices
cat("\n LDA Confusion Matrix:\n")
print(table(Predicted = pr.lda$class, Actual = test_iris$Species))

cat("\n NN Confusion Matrix:\n")
print(table(Predicted = pr.nn.class, Actual = test_iris$Species))

cat("\n QDA Confusion Matrix:\n")
print(table(Predicted = pr.qda$class, Actual = test_iris$Species))

# lda.fit = write code to fit LDA model to niris
# qda.fit = write code to fit QDA model to niris
```

The code below generates a grid on a square that contains all observations.

```{r}
xmin = min(niris$SLength)-1   #Obtaining the minimum, minus one to cover all observations
xmax = max(niris$SLength)+1   #Obtaining the maximum, plus one to cover all observations
ymin = min(niris$SWidth)-1   #Obtaining the minimum, minus one to cover all observations
ymax = max(niris$SWidth)+1   #Obtaining the maximum, plus one to cover all observations

many = 100        #Number of points per side of the grid
x_grid = seq(xmin,xmax,length.out = many) #Computing the grid edge
y_grid = seq(ymin,ymax,length.out = many) #Computing the grid edge
```


In the code below, we first plot the data points. Then we go through all grid points, predict their class, and draw the gridpoint coloured according to class. Edit the commented code to draw the classification regions:

```{r}

par(mfrow=c(1,2))
par(mar=c(0.1,0.1,2,0.1))
plot(X,pch=y-1,col=y+1,main="LDA",yaxt='n',xaxt='n')

for(j in 1:many){
  for(k in 1:many){
    new = data.frame(SLength=x_grid[j], SWidth = y_grid[k])
    # new.pred.lda = use predict() to predict the class of the observation "new" using LDA
    new.pred.lda = predict(lda.fit, new)
    curr_obs=as.numeric(new.pred.lda$class)
    points(x_grid[j],y_grid[k],col=curr_obs+1,pch=3,cex=0.5)  
  }
}

plot(X,pch=y-1,col=y+1,main="QDA",yaxt='n',xaxt='n')
for(j in 1:many){
  for(k in 1:many){
    new = data.frame(SLength=x_grid[j], SWidth = y_grid[k])
    # new.pred.qda = use predict() to predict the class of the observation "new" using QDA
    new.pred.qda = predict(qda.fit, new)
    curr_obs=as.numeric(new.pred.qda$class)
    points(x_grid[j],y_grid[k],col=curr_obs+1,pch=3,cex=0.5)   
  }
}
```

Principal Component Analysis and K-means

```{r}
data = scale(iris[,1:4])
print(cor(data))

p = princomp(x = data)
summary(p)
scores = p$scores[, 1:2]

biplot(p)

set.seed(50)
km = kmeans(p$scores[, 1:2], 3)

plot(scores,
     col  = km$cluster,
     pch  = 19,
     main = "K-means (K=3) on First Two PCs",
     xlab = "PC1", ylab = "PC2")
legend("topright",
       legend = paste("Cluster", 1:3),
       col    = 1:3, pch = 19)

# Plot coloured by species
plot(scores,
     col  = as.numeric(iris$Species),
     pch  = 19,
     main = "True Species on First Two PCs",
     xlab = "PC1", ylab = "PC2")
legend("topright",
       legend = levels(iris$Species),
       col    = 1:3, pch = 19)

```

Regulariseret regression

install.packages("ISLR")
install.packages("glmnet")
library(ISLR)
library(glmnet)

```{r}
set.seed(55)
data(College)

index_c = sample(1:nrow(College), round(0.7 * nrow(College)), replace = FALSE)
train_c = College[index_c, ]
test_c = College[-index_c, ]

OLS <- lm(Apps ~ ., data = train_c)
OLS_p <- predict(OLS, test_c)
OLS_mse <- mean((test_c$Apps - OLS_p)^2)
cat("OLS test MSE:", OLS_mse, "\n")

x_train <- model.matrix(Apps ~ ., train_c)[, -1]
y_train <- train_c$Apps
x_test  <- model.matrix(Apps ~ ., test_c)[, -1]
y_test  <- test_c$Apps

# Alpha er 0 da dette giver ridge regression (Elastic net)
cv_ridge   <- cv.glmnet(x_train, y_train, alpha = 0, standardize = TRUE)
best_lam_r <- cv_ridge$lambda.min
cat("Ridge best lambda:", best_lam_r, "\n")

ridge_pred <- predict(cv_ridge, s = best_lam_r, newx = x_test)
ridge_mse  <- mean((y_test - ridge_pred)^2)
cat("Ridge test MSE:", ridge_mse, "\n")

plot(cv_ridge)

# Alpha er 1 da dette giver lasso regression (Elastic net)
cv_lasso   <- cv.glmnet(x_train, y_train, alpha = 1, standardize = TRUE)
best_lam_l <- cv_lasso$lambda.min
cat("Lasso best lambda:", best_lam_l, "\n")

lasso_pred <- predict(cv_lasso, s = best_lam_l, newx = x_test)
lasso_mse  <- mean((y_test - lasso_pred)^2)
cat("Lasso test MSE:", lasso_mse, "\n")

lasso_coef    <- predict(cv_lasso, s = best_lam_l, type = "coefficients")
nonzero_count <- sum(lasso_coef != 0) - 1 
cat("Lasso non-zero coefficients:", nonzero_count, "\n")

plot(cv_lasso)

results <- data.frame(
  Method = c("OLS", "Ridge", "Lasso"),
  Test_MSE = c(OLS_mse, ridge_mse, lasso_mse)
)
print(results)

```

Bagging and Boosting

install.packages(c("rpart", "rpart.plot", "randomForest", "gbm"))
library(randomForest)
library(gbm)

```{r, ehco=FALSE}
# Data handling
day = read.csv("H:\\AAU - Uni\\Statistics Learning - Data mining\\Bike-Sharing-Dataset\\day.csv")

day$season     <- factor(day$season)
day$holiday    <- factor(day$holiday)
day$workingday <- factor(day$workingday)
day$weathersit <- factor(day$weathersit)

set.seed(42)
index_b1   <- sample(nrow(day), 0.8 * nrow(day))
train <- day[index_b1, ]
test  <- day[-index_b1, ]

# Bagging
p <- 7

bag_fit <- randomForest(
  cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
  data       = train,
  mtry       = p,       
  ntree      = 500,
  importance = TRUE
)

varImpPlot(bag_fit, main = "Bagging — variable importance")
importance(bag_fit)   

preds_bag <- predict(bag_fit, newdata = test)
cat("\n Bagging MSE: \n")
mean((test$cnt - preds_bag)^2)

# Boosting
depths <- c(1, 2, 3, 4, 6)
mse_boost <- numeric(length(depths))

for (i in seq_along(depths)) {
  boost <- gbm(
    cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
    data              = train,
    distribution      = "gaussian",   # regression
    n.trees           = 1000,
    interaction.depth = depths[i],
    shrinkage         = 0.01,
    cv.folds          = 5,            
    verbose           = FALSE
  )
  best_iter <- gbm.perf(boost, method = "cv", plot.it = FALSE)
  preds     <- predict(boost, newdata = test, n.trees = best_iter)
  mse_boost[i] <- mean((test$cnt - preds)^2)
}

data.frame(depth = depths, MSE = round(mse_boost, 1))

best_depth <- depths[which.min(mse_boost)]
boost_best <- gbm(
  cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
  data              = train,
  distribution      = "gaussian",
  n.trees           = 1000,
  interaction.depth = best_depth,
  shrinkage         = 0.01,
  cv.folds          = 5,
  verbose           = FALSE
)
best_iter <- gbm.perf(boost_best, method = "cv", plot.it = FALSE)
summary(boost_best, n.trees = best_iter)  
```
Bagging and Boosting MSE over B trees

```{r}
# Data handling
day = read.csv("H:\\AAU - Uni\\Statistics Learning - Data mining\\Bike-Sharing-Dataset\\day.csv")

day$season     <- factor(day$season)
day$holiday    <- factor(day$holiday)
day$workingday <- factor(day$workingday)
day$weathersit <- factor(day$weathersit)

set.seed(42)
index_b2   <- sample(nrow(day), 0.8 * nrow(day))
train <- day[index_b2, ]
test  <- day[-index_b2, ]

# Bagging
bag_fit <- randomForest(
  cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
  data       = train,
  mtry       = 7,  
  ntree      = 5000,
  importance = TRUE
)

bag_mse <- sapply(1:5000, function(n) {
  preds <- predict(bag_fit, newdata = test, predict.all = TRUE)$individual[, 1:n, drop = FALSE]
  row_means <- rowMeans(preds)
  mean((test$cnt - row_means)^2)
})


# --- Boosting: depth 1 ---
boost_d1 <- gbm(
  cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
  data              = train,
  distribution      = "gaussian",
  n.trees           = 5000,
  interaction.depth = 1,
  shrinkage         = 0.01,
  verbose           = FALSE
)

boost_d1_mse <- sapply(1:5000, function(n) {
  preds <- predict(boost_d1, newdata = test, n.trees = n)
  mean((test$cnt - preds)^2)
})

# --- Boosting: depth 2 ---
boost_d2 <- gbm(
  cnt ~ atemp + hum + windspeed + season + holiday + workingday + weathersit,
  data              = train,
  distribution      = "gaussian",
  n.trees           = 5000,
  interaction.depth = 2,
  shrinkage         = 0.01,
  verbose           = FALSE
)

boost_d2_mse <- sapply(1:5000, function(n) {
  preds <- predict(boost_d2, newdata = test, n.trees = n)
  mean((test$cnt - preds)^2)
})

# --- Plot ---
plot(1:5000, bag_mse,
     type = "l", col = "cyan3", lwd = 2,
     ylim = range(c(bag_mse, boost_d1_mse, boost_d2_mse)),
     xlab = "Number of Trees", ylab = "Test MSE",
     main = "Test MSE vs Number of Trees")

lines(1:5000, boost_d1_mse, col = "gold3",  lwd = 2)
lines(1:5000, boost_d2_mse, col = "orange", lwd = 2)

legend("topright",
       legend = c("Boosting: d=1", "Boosting: d=2", paste0("Bagging")),
       col    = c("gold3", "orange", "cyan3"),
       lwd    = 2)
```

Support Vector Machine

install.packages('e1071')
install.packages('caret')
library(e1071)
library(caret)

```{r}
set.seed(70)

data(iris)
index_s <- sample(1:nrow(iris), 0.5 * nrow(iris))
iris_train <- iris[index_s, ]
iris_test <- iris[-index_s, ]

# Linear SVM
model_linear <- svm(Species ~ ., data = iris_train, kernel = "linear", cost = 0.1, scale = TRUE)

tuned_linear <- tune(svm, Species ~ ., data = iris_train, 
                     ranges = list(epsilon = seq(0, 1, 0.1), 
                                   cost = 2^(-3:7)))
plot(tuned_linear)
best_linear <- tuned_linear$best.model

# Confusion matrices
predicted_ini <- predict(model_linear, iris_test[, -5])
predicted_best <- predict(best_linear, iris_test[, -5])
confusionMatrix(table(predicted_ini, iris_test$Species))
confusionMatrix(table(predicted_best, iris_test$Species))

# Decision boundary plot
plot(model_linear, iris_train,
     Petal.Width ~ Petal.Length,
     slice = list(Sepal.Length = mean(iris$Sepal.Length),
                  Sepal.Width  = mean(iris$Sepal.Width)),
     main = "SVM classification plot (linear)")
plot(best_linear, iris_train,
     Petal.Width ~ Petal.Length,
     slice = list(Sepal.Length = mean(iris$Sepal.Length),
                  Sepal.Width  = mean(iris$Sepal.Width)),
     main = "Best SVM classification plot (linear)")

# Radial SVM
model_radial <- svm(Species ~ ., data = iris_train,
                    kernel = "radial", cost = 1, gamma = 0.5,
                    scale = TRUE)

tuned_radial <- tune(svm, Species ~ ., data = iris_train,
                     kernel = "radial",
                     ranges = list(cost  = 2^(-3:6),
                                  gamma = 10^(-3:1)))
plot(tuned_radial)
best_radial <- tuned_radial$best.model

# Confusion matrices
predicted_ini <- predict(model_radial, iris_test[, -5])
predicted_best <- predict(best_radial, iris_test[, -5])
confusionMatrix(table(predicted_ini, iris_test$Species))
confusionMatrix(table(predicted_best, iris_test$Species))

# Decision boundary
plot(model_radial, iris_train,
     Petal.Width ~ Petal.Length,
     slice = list(Sepal.Length = mean(iris$Sepal.Length),
                  Sepal.Width  = mean(iris$Sepal.Width)),
     main = "SVM classification plot (radial)")
plot(best_radial, iris_train,
     Petal.Width ~ Petal.Length,
     slice = list(Sepal.Length = mean(iris$Sepal.Length),
                  Sepal.Width  = mean(iris$Sepal.Width)),
     main = "Best SVM classification plot (radial)")
```
NN compared to LDA and QDA using all features

```{r}
# Data handling
set.seed(120)
data(iris)

X = iris[,1:4]
niris = data.frame(Species=y,
                   SLength = X[,1], SWidth  = X[,2],
                   PLength = X[,3], PWidth  = X[,4])

niris2 = data.frame(SLength = X[,1], SWidth  = X[,2],
                    PLength = X[,3], PWidth  = X[,4])
niris2$setosa     <- as.integer(iris[,5] == "setosa")
niris2$versicolor <- as.integer(iris[,5] == "versicolor")
niris2$virginica  <- as.integer(iris[,5] == "virginica")

index_da = sample(1:nrow(niris), round(0.5 * nrow(niris)), replace=FALSE)
train_iris = niris[index_da, ]
test_iris = niris[-index_da, ]
test_iris$Species  <- factor(test_iris$Species,  levels=1:3, labels=c("setosa","versicolor","virginica"))
train_iris$Species <- factor(train_iris$Species, levels=1:3, labels=c("setosa","versicolor","virginica"))
index_nn = sample(1:nrow(niris2), round(0.5 * nrow(niris2)), replace = FALSE)
train_iris2 = niris2[index_nn, ]
test_iris2 = niris2[-index_nn, ]

# Model fit and predictions
lda.fit = lda(Species ~ SLength + SWidth + PLength + PWidth, data=train_iris)
qda.fit = qda(Species ~ SLength + SWidth + PLength + PWidth, data=train_iris)
nn = neuralnet("setosa + versicolor + virginica ~ SLength + SWidth + PLength + PWidth",
               data=train_iris2, hidden=c(4, 2), linear.output=FALSE, act.fct="logistic")
plot(nn)
pr.nn = compute(nn, test_iris2[, c("SLength", "SWidth", "PLength", "PWidth")])

pr.lda = predict(lda.fit, test_iris)
pr.qda = predict(qda.fit, test_iris)

pr.nn.mat = pr.nn$net.result
colnames(pr.nn.mat) = c("setosa", "versicolor", "virginica")

pr.nn.class = apply(pr.nn.mat, 1, which.max)
pr.nn.class = factor(pr.nn.class, levels=1:3, labels = c("setosa", "versicolor", "virginica"))

# Confusion matrices
cat("\n LDA Confusion Matrix:\n")
print(table(Predicted = pr.lda$class, Actual = test_iris$Species))

cat("\n NN Confusion Matrix:\n")
print(table(Predicted = pr.nn.class, Actual = test_iris$Species))

cat("\n QDA Confusion Matrix:\n")
print(table(Predicted = pr.qda$class, Actual = test_iris$Species))

# Plot setup
xmin = min(niris$PLength) - 1
xmax = max(niris$PLength) + 1
ymin = min(niris$PWidth)  - 1
ymax = max(niris$PWidth)  + 1

many    = 100
x_grid  = seq(xmin, xmax, length.out = many)
y_grid  = seq(ymin, ymax, length.out = many)

mean_SLength = mean(niris$SLength)
mean_SWidth  = mean(niris$SWidth)

par(mfrow = c(1, 2))
par(mar  = c(0.1, 0.1, 2, 0.1))

# Plot of LDA
plot(niris$PLength, niris$PWidth,
     pch = y - 1, col = y + 1,
     main = "LDA", xaxt = 'n', yaxt = 'n')

for (j in 1:many) {
  for (k in 1:many) {
    new = data.frame(SLength = mean_SLength,
                     SWidth  = mean_SWidth,
                     PLength = x_grid[j],
                     PWidth  = y_grid[k])
    new.pred.lda = predict(lda.fit, new)
    curr_obs = as.numeric(new.pred.lda$class)
    points(x_grid[j], y_grid[k], col = curr_obs + 1, pch = 3, cex = 0.5)
  }
}

# Plot of QDA
plot(niris$PLength, niris$PWidth,
     pch = y - 1, col = y + 1,
     main = "QDA", xaxt = 'n', yaxt = 'n')

for (j in 1:many) {
  for (k in 1:many) {
    new = data.frame(SLength = mean_SLength,
                     SWidth  = mean_SWidth,
                     PLength = x_grid[j],
                     PWidth  = y_grid[k])
    new.pred.qda = predict(qda.fit, new)
    curr_obs = as.numeric(new.pred.qda$class)
    points(x_grid[j], y_grid[k], col = curr_obs + 1, pch = 3, cex = 0.5)
  }
}

# Plot of NN
grid_df = expand.grid(PLength = x_grid,
                      PWidth  = y_grid)
grid_df$SLength = mean_SLength
grid_df$SWidth  = mean_SWidth

pr_grid      = compute(nn, grid_df[, c("SLength", "SWidth", "PLength", "PWidth")])
grid_df$class = apply(pr_grid$net.result, 1, which.max)

plot(niris$PLength, niris$PWidth,
     pch = y - 1, col = y + 1,
     main = "NN", xaxt = 'n', yaxt = 'n')

points(grid_df$PLength, grid_df$PWidth,
       col = grid_df$class + 1, pch = 3, cex = 0.5)

```












