---
title: "Linear Classification: Heart Disease"
author: "Data Mining"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

In this exercise we are going to analyze coronary heart disease determinants. We use data taken from a larger
dataset, described in  Rousseauw et al, 1983, South African Medical Journal. It contains measurements for several variables of males in a heart-disease high-risk region of the Western Cape, South Africa. These variables are, among others,

Variable    | Description 
------------| -------------
*tobacco*		|cumulative tobacco (kg)
*ldl*	    	|low density lipoprotein cholesterol
*famhist*		|family history of heart disease (Present, Absent)
*age*		    |age at onset
*chd*		    |response, coronary heart disease

The data is available online and can be loaded directly into R.

```{r}
heart = read.table("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/SAheart.data",
                 sep=",",head=T,row.names=1)
```

We use *attach()* so that we can refer to the names of the variables directly.

```{r}
attach(heart)
```

The response variable is the presence of coronary heart disease. As explanatory variables we will use *tobacco*, *ldl*, *famhist*, and *age*.
As such, we are in a binary classification setting. 

Below, we will fit a logistic regression and an LDA model to the data. You can find examples of how to implement the different model in ISLR Chapter 4.7.

## Logistic Regression

- Fit a logistic regression model *chd* depends on the variables *tobacco*, *ldl*, *famhist*, and *age*. You may use the *glm()* function with *family=binomial*. To see the results, use the *summary()* command.
- Estimate and interpret the odds-ratio for *age* and *famhist*. 

## Linear Discriminant Analysis

- Fit an LDA model to the data set using the *lda()* function from the *MASS* package.
- Compare the parameter estimates to those from logistic regression and comment on the results.

## Prediction

- Use the models fitted above to predict chd for a person aged 43 who has smoked a total of 3kg tobacco, has a family history of coronary heart disease and a measured ldl level of 5. (You can use the *predict()* function. Note that for logistic regression, this gives the class probabilities, while for LDA, this gives the predicted class. The class probabilities for can be obtained by *predict()$posterior*.)  


```{r}
library(MASS)

prediction_data = data.frame(age = 43,
                             tobacco = 3,
                             famhist = "Present",
                             ldl = 5)

lda.fit = lda(chd ~ tobacco + ldl + famhist + age, data=heart)
pred = predict(lda.fit, prediction_data)
print(pred)


```






